### 2022-02-17 pre/pre.tex
more

\chapter{Preliminaries}
\setheader{Preliminaries}
\label{ch:preliminatires}

>>>>

\chapter{Preliminaries}
\setheader{Preliminaries}
\label{ch:pre}

\begin{abstract}
    In this chapter we lay out the basic prerequisites for the remaining
    chapters. We begin by defining three logics that we work with in this
    thesis: propositional logic, monomorphic first-order logic and simply typed
    higher-order logic. Then, we explain the clausal structure which is the
    backbone of many calculi for automated provers. We further describe the
    superposition calculus and finish with description of the process superposition
    provers use to actually perform the inferences.
\end{abstract}

\newpage

Our work mostly concerns retrofitting first-order provers to support
higher-order concepts. But part of our work also concerns integrating
successful approaches for propositional logic in first-order
provers. We begin by defining all three logics, in the increasing
order of expressivity.

\section{Propositional Logic}

Basic formulas of propositional logic syntax are propositional variables
$\cst{p}, \cst{q}, \ldots$ and constants $\top$ and $\bot$. More complex
formulas are built inductively using logical connectives $\neg, \land, \lor,
\imp, \lequiv$: if $\phi$ and $\psi$ are formulas, then $\neg \phi, \phi \land
\psi, \phi \lor \psi, \phi \imp \psi, \phi \lequiv \psi$ are formulas as well.

To interpret the formulas, propositional variables are assigned 0 (false) or 1
(true), constants $\top$ and $\bot$ are assigned 1 and 0 respectively and the
formula is interpreted using the rules for each connective
\cite[Sect.~1.4]{hr-00-logic-in-cs}. As there are finitely many propositional
variables in a formula, trying all (finitely many) possible assignments
describes an algorithm to decide satisfiability (existence of satisfiable
assignment) or validity (if all assignments are satisfiable).

However, this simple approach is prohibitively expensive and modern tools that
decide propositional satisfiability problem (SAT solvers) use more advanced
approaches such as CDCL calculus \cite{mss-96-cdcl}. Modern SAT solvers also
heavily preprocess the problem and continue simplifying it during proving
process.

\section{First-Order Logic}

First-order logic increases the expressivity by allowing quantification over
objects and has more complicated formula structure. There are many flavors of
first-order logic, but in this thesis we consider monomorphic first-order logic
with equality.

We distinguish a set of base types $T$ which is required to have the Boolean
type $o$, and a set of symbols $\Sigma$. To each symbol $\cst{f} \in \Sigma$ a
tuple $(\tau_1, \ldots, \tau_{n}, \tau), n \geq 0$ of types is assigned, written
$\cst{f} : (\tau_1, \ldots, \tau_{n}) \rightarrow \tau$. We say that $(\tau_1,
\ldots, \tau_{n})$ are argument types, $\tau$ is the return type, and that $n$
is the arity of the symbol $\cst{f}$. If return type of $\cst{f}$ is $o$ we call
$\cst{f}$ a predicate symbol, otherwise we call it a function symbol. It is
required that argument types are not Boolean.

Terms are the basic building blocks of first-order logic. Variables
$x,y,z,\ldots$, assigned types $\tau \in T$, are terms. If $t_1,\ldots,t_n$ are
terms of types $\tau_1, \ldots, \tau_n$, respectively, and $\cst{f} : (\tau_1,
\ldots, \tau_{n}) \rightarrow \tau$, then $\cst{f}(t_1, \ldots, t_n)$ is a term
of type $\tau$. If $n=0$, we drop parentheses and write $\cst{f}$. We also
abbreviate $\cst{f}(t_1, \ldots, t_n)$ to $\cst{f}(\tuplen{t})$.

Terms are used to build atoms. Atom is either a term $t : o$ or an equation $s
\eq t$ where terms $t$ and and $s$ are of the same type. Atoms are combined
using logical connectives just like in propositional logic. Additionally,
first-order formulas are built using quantifiers: if $\phi$ is a formula and $x$
is a variable then $\forall x.\, \phi$, as well as $\exists x.\, \phi$ are
formulas. The first formula is interpreted as ``$\phi$ must hold for all terms
replaced for $x$'' and the second one as ``there is a replacement for $x$ such
that $\phi$ holds''. As soon as there is a single functional symbol with arity
greater than 0, there are infinitely many terms that can be substituted for a
free variable. It is therefore obvious that the propositional technique for
deciding satisfiability does not work in the first-order case.

Substitutions $\sigma, \varrho, \ldots$ are total mappings from variables to
terms of the same type. Substitutions map only finitely many variables to terms
other than the variable itself. This is denoted as $\{ x_1 \mapsto t_1, \ldots,
x_n \mapsto t_n \}$ where $x_i$ are variables that are not mapped to itself. 
Applying a substitution $\sigma$ to a term $t$ results in replacing all mapped variables 
by the corresponding terms and it is denoted.
\section{Higher-Order Logic}
\section{Clausal Forms}
\section{Superposition}
\subsection{Term Order and Selection}
\subsection{Unification}
\subsection{The Calculus}
\subsection{Redundancy Criterion}
\subsection{Saturation Procedure}
